D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-19 ootavacuous.tex
Edits for Appendix D5
The text for Appendix D5 was wrong in several respects.  The program
doesn't have UB, and the OOTA-like result of the optimization proves
that the optimization was invalid.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\subsection{OOTA From Undefined-Behavior Optimizations}
\label{app:OOTA From Undefined-Behavior Optimizations}

This section departs from one of the rules for this paper by considering
an example having undefined behavior.

Many compilers carry out optimizations based on the assumption
that UB cannot happen, and these optimizations can result in OOTA cycles,
with a simple example shown in
Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
\caption{OOTA Due To Divide-By-Zero UB}
\label{lst:OOTA Due To Divide-By-Zero UB}
\end{listing}

On line~14 of this listing, the value of \co{r2} must be either zero
or one due to the comparison on line~13.
If it is the value one, then there is a divide-by-zero error, which
enables the compiler to assume that the value of \co{r1} is
always zero.
This in turn allows the compiler to behave as if the expression
stored to \co{y} on line~14 was the constant 1, at which point there
is no semantic dependency from the load from \co{x} on line~13 to the
store to \co{y} on line~14.

The compiler is then within its rights to act as if lines~13 and~14
were interchanged.
At which point, the order of execution could be as follows:

\begin{enumerate}
\item	Line~14 stores the constant 1 to \co{y}.
\item	Line~8 loads the constant 1 from \co{y}, setting \co{r1} to 1.
\item	Line~9 stores the constant 1 to \co{x}.
\item	Line~8 loads the constant 1 from \co{x}, setting \co{r2} to 1.
\end{enumerate}

Please note that the OOTA-like cycle called out in the \co{exists}
clause can happen even though the program never actually executes any
code that would have resulted in division-by-zero UB.
It is tempting to consider using techniques
% Appendix~\ref{app:Aside on Undefined Behavior}
that prevent back-propagation of UB, but these are ineffective in this
case because the compiler can detect and exploit UB when
considering line~14 in isolation.
Once the constant 1 is substituted for line~14's expression, the
hardware is free to reorder lines~13 and~14 even if a compiler-based
C++ implementation refrains from doing so.
Note that this OOTA-like cycle can appear even if all of the atomic
operations in
Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
are converted to volatile.

Therefore, this paper focuses on UB-free programs, rare though these
beasts might be among large-scale C++ programs running in production.

>>>>

\subsection{Undefined Behavior and Unwise Optimization}
\label{app:Undefined Behavior and Unwise Optimization}

Many compilers carry out optimizations based on the assumption
that undefined behavior (UB) cannot happen,
and if not performed carefully, these optimizations may be invalid
because of OOTA-like interactions.
A simple example is shown in
Listing~\ref{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
\caption{OOTA-Like Behavior Due To Divide-By-Zero UB}
\label{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}
\end{listing}

On line~14 of this listing, the computation \co{1 / (r2 <= 0)}
will yield one if \co{r2} is not positive and will result in UB
otherwise (a divide-by-zero error).
It may not be obvious, but the program is in fact UB-free---there are
no abstract executions in which \co{r2} is greater than zero.
To see why not, consider that in any such execution
\co{P1()} would not store anything to \co{y};
it certainly would not store a positive value.
This means that the value loaded on line~8 would have to be \co{y}'s
initial value of zero, and so would the value stored to \co{x} on
line~9, and hence it would not be possible for the load on line~13
to get a positive value for \co{r2} after all.

Even for a compiler analyzing \co{P1()} in isolation,
it is clear that either line~14 will store one to \co{y} or
else the division by zero will cause UB.
This allows the compiler to assume that the store will always
take place, on the grounds that UB cannot happen,
and so the compiler may optimize the comparison and division
by replacing the whole expression with a constant \co{1}.
There will then be no dependency from the load on line~13
to the store and thus nothing forcing the store to execute
after the load.

When the resulting program runs on a weakly ordered architecture,
the execution could go as follows:
\begin{enumerate}
\item	Line~14 executes out of order, storing the constant \co{1} to \co{y}.
\item	Line~8 loads one from \co{y}, setting \co{r1} to one.
\item	Line~9 stores one to \co{x}.
\item	Line~13 loads one from \co{x}, setting \co{r2} to one.
\end{enumerate}
The final result would satisfy the \co{exists} clause,
exhibiting observable behavior (\co{r2} is one at the end)
that no abstract execution of the original program could produce.
(This could happen even if all of the atomic objects in
Listing~\ref{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}
were converted to volatile.)

This result shows that the proposed optimization was invalid.
To prevent the unwanted behavior, the compiler would have to take an
extra step to force the optimized store on line~14 to be ordered after
the load on line~13, possibly by changing the load to a load-acquire
or changing the store to a store-release.

In other words, the compiler would have to preserve the dependency
ordering that the UB-based optimization would otherwise destroy.
It is tempting to consider using techniques
% Appendix~\ref{app:Aside on Undefined Behavior}
that prevent back-propagation of UB, but these are ineffective in this
case because the compiler can detect and exploit UB when
considering line~14 in isolation.
